---
title: "Batting Model"
output:   
  html_document: 
    code_folding: hide
    toc: true
    toc_float: true
    df_print: kable

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rvest)
library(readxl)
library(scales)
library(patchwork)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%",
  dpi = 300,
  warning = FALSE,
  message = FALSE
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

salaries = read_excel(
  "data/MLB-Salaries 2000-24.xlsx", 
  sheet = "2022.xls",
  skip = 1) |>
  select(1:4) |>
  rename(position = "Pos'n",
         salary_2022 = "2022.0",
         service_time_yrs = "MLS",
         name = "Player") %>%
  mutate(name = str_split(name, ","),
         name = map(name, rev),       
         name = map_chr(name, str_c, collapse = " "),
         name = str_trim(name),
         name = str_replace_all(name, "[*#\\.,]", ""),
         junior = str_detect(name, "Jr"),
         name = if_else(junior == TRUE, str_remove(name, " Jr"), name),
         name = if_else(junior == TRUE, str_c(name, " Jr"), name),
         name = stringi::stri_trans_general(name,id = "Latin-ASCII")) |>
  select(-junior) %>%
  mutate(simple_position = str_split_i(position, "-", 1),
         simple_position = fct_relevel(simple_position,
                                       c("c", "1b", "2b", "3b", "ss",
                                         "lf", "cf", "rf", "inf", "of",
                                         "dh", "rhp", "lhp")),
         service_time_floor = floor(service_time_yrs),
         experience_group = case_when(service_time_floor <= 2 ~ "0-2 years",
                                      service_time_floor >= 3 ~ "3+ years"))

batting = read_delim("data/2022 MLB Player Stats - Batting.csv", delim = ";",
                     locale = locale(encoding = "latin1")) |>
  janitor::clean_names() %>%
  group_by(name) |>
  mutate(name_count = n(),
         keep_row = case_when(name_count == 1 ~ TRUE,
                              name_count > 1 & tm == "TOT" ~ TRUE,
                              .default = FALSE)) |>
  filter(keep_row == TRUE) %>%
  mutate(name_count = n(),
         keep_row = case_when(name_count == 1 ~ TRUE,
                              name_count > 1 & lg == "MLB" ~ TRUE,
                              .default = FALSE)) |>
  ungroup() |>
  filter(keep_row == TRUE) %>%
  mutate(name = str_split(name, "\\s+"),
         name = map_chr(name, str_c, collapse = " "),
         name = str_trim(name),
         name = str_replace_all(name, "[*#\\.]", ""),
         name = stringi::stri_trans_general(name,id = "Latin-ASCII"))

merged_batting <- salaries |>
  mutate(simple_position = str_split_i(position, "-", 1)) |>
  filter(!simple_position %in% c("rhp","lhp")) |>
  inner_join(batting, by = "name") |>
  filter(pa >= 100)

salary_not_in_batting <- anti_join(salaries, batting, by = "name")
batting_not_in_salary <- anti_join(batting, salaries, by = "name")
```

```{r}
lm_batting = lm(log10(salary_2022) ~ ops_2 + experience_group,
                 data = merged_batting)

broom::tidy(lm_batting)

broom::glance(lm_batting)
```

```{r}

plot(residuals(lm_batting) ~ fitted.values(lm_batting), 
     main = "Residuals vs. Fitted Values Plot for Batting Model",
     xlab = "Fitted Values",
     ylab = "Residuals")
abline(h = 0)

```

```{r}

qqnorm(residuals(lm_batting),
       main = "Normal Q-Q Plot for Batting Model")
qqline(residuals(lm_batting))

```

We decided to log-transform the salary variable because it is heavily right-skewed, and decided to use $log_{10}$ instead of $ln$ for the sake of interpretability. We included OPS+, the "all-in-one" measure of hitter performance as well as their number of years of experience.

From our model, we can see that batters with more experience have higher salaries, even adjusting for performance. At the same time, adjusting for service time, batters with higher OPS+ (that is, batters who perform better) have significantly higher salaries.

Based on this model, which player provided the best value for money? Who was most cost effective? That would be the player with a high  OPS+ and the lowest residual, indicating that their actual salary was lower than predicted based on their performance.

### Top 5 Batters by Value

```{r}
merged_batting |>
  modelr::add_predictions(lm_batting) |>
  modelr::add_residuals(lm_batting) |>
  arrange(resid) |>
  select(name, salary_2022, service_time_floor, experience_group, ops_2, pred, resid ) |>
  mutate(expected_salary = 10^pred) |>
  slice(1:5) 
```

### Top 10 Batters by Salary

```{r}
merged_batting |>
  modelr::add_predictions(lm_batting) |>
  modelr::add_residuals(lm_batting) |>
  arrange(desc(salary_2022)) |>
  select(name, salary_2022, service_time_floor,experience_group, ops_2, pred, resid ) |>
  mutate(expected_salary = 10^pred) |>
  slice(1:10) 
```

It looks like in 2022, the most cost-effective player was Albert Pujols, who made $2.5 million in his 21st season with an OPS+ of 151.
