---
title: "Batting EDA"
output:   
  html_document: 
    code_folding: hide
    toc: true
    toc_float: true
    df_print: kable

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rvest)
library(readxl)
library(scales)
library(patchwork)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%",
  dpi = 300,
  warning = FALSE,
  message = FALSE
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

batting = read_delim("data/2022 MLB Player Stats - Batting.csv", delim = ";",
                     locale = locale(encoding = "latin1")) |>
  janitor::clean_names() %>%
  group_by(name) |>
  mutate(name_count = n(),
         keep_row = case_when(name_count == 1 ~ TRUE,
                              name_count > 1 & tm == "TOT" ~ TRUE,
                              .default = FALSE)) |>
  filter(keep_row == TRUE) %>%
  mutate(name_count = n(),
         keep_row = case_when(name_count == 1 ~ TRUE,
                              name_count > 1 & lg == "MLB" ~ TRUE,
                              .default = FALSE)) |>
  ungroup() |>
  filter(keep_row == TRUE) %>%
  mutate(name = str_split(name, "\\s+"),
         name = map_chr(name, str_c, collapse = " "),
         name = str_trim(name),
         name = str_replace_all(name, "[*#\\.]", ""),
         name = stringi::stri_trans_general(name,id = "Latin-ASCII"))

merged_batting <- salaries |>
  mutate(simple_position = str_split_i(position, "-", 1)) |>
  filter(!simple_position %in% c("rhp","lhp")) |>
  inner_join(batting, by = "name") |>
  filter(pa >= 100)

salary_not_in_batting <- anti_join(salaries, batting, by = "name")
batting_not_in_salary <- anti_join(batting, salaries, by = "name")
```

Many of the counting statistics for batting are right skewed, as most players are not hitting multiple home runs in the season. Some of the advanced statistics such as OPS+ are more normally distributed. An important idea to keep in mind is that many of these collected statistics are strongly correlated with each other as the advanced statistics are often a formula taking input from the counting statistics. Looking at the relation with salary, it seems that there may be an association between salary and home runs as well as OPS+. 

```{r}
# Chances at Bat by Home Runs
merged_batting %>%
  ggplot(aes(x = ab,y= hr)) +
  geom_point(aes(color = ops)) +
  geom_smooth(method = lm, se = FALSE, color = "firebrick") +
  labs(title = "Who is the most 'effective' at the plate?",
       x = "Chances at Bat",
       y = "Number of Home Runs") +
  scale_color_continuous("OPS") +
  theme_bw()
```

```{r}
# Corrs for Batting Data
merged_batting |>
  select(hr, ba, ops) |>
  GGally::ggpairs()
```

```{r}
# Distributions for Used Statistics
yrsCol = merged_batting |>
  count(experience_group) |>
  ggplot(aes(x = experience_group, y=n)) + 
  geom_col(fill = 'blue', col = 'black') +
  labs(x="Years of Experience",y="") +
  theme_bw()
hrsHist = merged_batting |>
  ggplot(aes(x = hr)) + 
  geom_histogram(fill = 'purple', col = 'black') +
  labs(x="Home Runs",y="") +
  theme_bw()
basHist = merged_batting |>
  ggplot(aes(x = ba)) + 
  geom_histogram(fill = 'yellow', col = 'black') +
  labs(x="Batting Averages",y="") +
  theme_bw()
opsHist = merged_batting |>
  ggplot(aes(x = ops_2)) + 
  geom_histogram(fill = 'lightgreen', col = 'black') +
  labs(x="OPS+",y="") +
  theme_bw()

dist_patch = (yrsCol+hrsHist)/(basHist+opsHist)

dist_patch + plot_annotation(title = "Distribution of Important Batting Statistics")
```

```{r}
# Scatter for Used Statistics
yrs = merged_batting |>
  ggplot(aes(x = experience_group, y = salary_2022)) + 
  geom_boxplot(col = 'blue') +
  labs(x="Years of Experience",y="") +
  scale_y_continuous(labels = label_comma()) +
  theme_bw()
hrs = merged_batting |>
  ggplot(aes(x = hr, y = salary_2022)) + 
  geom_point(col = 'purple') +
  guides(y="none") +
  labs(x="Home Runs",y="") +
  theme_bw()
bas = merged_batting |>
  ggplot(aes(x = ba, y = salary_2022)) + 
  geom_point(col = 'yellow') +
  labs(x="Batting Averages",y="") +
  scale_y_continuous(labels = label_comma()) +
  theme_bw()
ops = merged_batting |>
  ggplot(aes(x = ops_2, y = salary_2022)) + 
  geom_point(col = 'lightgreen') +
  guides(y="none") +
  labs(x="OPS+",y="") +
  theme_bw()

scatter_patch = (yrs+hrs)/(bas+ops)

scatter_patch + plot_annotation(title = "Scatterplot of Batting Statistics and Salary")
```
