---
title: "Pitching EDA"
output:   
  html_document: 
    code_folding: hide
    toc: true
    toc_float: true
    df_print: kable

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rvest)
library(readxl)
library(scales)
library(patchwork)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%",
  dpi = 300,
  warning = FALSE,
  message = FALSE
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

pitching = read_delim("data/2022 MLB Player Stats - Pitching.csv", delim = ";",
                     locale = locale(encoding = "latin1")) |>
  janitor::clean_names() %>%
  group_by(name) |>
  mutate(name_count = n(),
         keep_row = case_when(name_count == 1 ~ TRUE,
                              name_count > 1 & tm == "TOT" ~ TRUE,
                              .default = FALSE)) |>
  filter(keep_row == TRUE) %>%
  mutate(name_count = n(),
         keep_row = case_when(name_count == 1 ~ TRUE,
                              name_count > 1 & lg == "MLB" ~ TRUE,
                              .default = FALSE)) |>
  ungroup() |>
  filter(keep_row == TRUE) %>%
  mutate(name = str_split(name, "\\s+"),
         name = map_chr(name, str_c, collapse = " "),
         name = str_trim(name),
         name = str_replace_all(name, "[*#\\.]", ""),
         name = stringi::stri_trans_general(name,id = "Latin-ASCII"))

merged_pitching <- inner_join(salaries, pitching, by = "name") |>
  filter(ip >= 20) |>
  separate(ip, into = c("ip", "ip_dec"), remove = FALSE, convert = TRUE) |>
  mutate(ip_dec_333 = ifelse(is.na(ip_dec), 0, ip_dec * 333)  ,
         ip_total = paste(ip, ip_dec_333, sep = "."),
         ip_total = as.numeric(ip_total)) |>
  select(-ip, -ip_dec, -ip_dec_333) |>
  separate(position, into = c("hand", "pitcher_type")) |>
  mutate(pitcher_type = ifelse(is.na(pitcher_type), "r", pitcher_type)) |>
  filter(hand == "lhp" | hand == "rhp")

salary_not_in_pitching <- anti_join(salaries, pitching, by = "name")
pitching_not_in_salary <- anti_join(pitching, salaries, by = "name")
```

```{r}

ip_plot = merged_pitching |> 
  ggplot(aes(x = ip_total)) + 
  geom_histogram(binwidth = 5, fill = "red", col = "black") +
  labs(
       x = "Innings Pitched",
       y = "") +
  theme_bw()

era_plot = merged_pitching |>
  ggplot(aes(x = era_2)) + 
  geom_histogram(fill = "lightblue", col = "black") +
    labs(
       x = "ERA+",
       y = "") +
  theme_bw()

bb_plot = merged_pitching |>
  ggplot(aes(x = bb9)) + 
  geom_histogram(fill = "darkgreen", col = "black") +
    labs(
       x = "Walks per 9 Innings (BB/9)",
       y = "") +
  theme_bw()

so_plot = merged_pitching |>
  ggplot(aes(x = so9)) + 
  geom_histogram(fill = "orange", col = "black") +
    labs(
       x = "Strikeouts per 9 Innings (SO/9)",
       y = "") +
  guides() +
  theme_bw()

hist_pitch = (ip_plot + era_plot) / (bb_plot + so_plot)
hist_pitch + plot_annotation(title = "Histograms of Pitching Statistics")

```

The distribution of innings pitched is quite right-skewed, with a peak around 70 innings pitched. The distribution of ERA+ is right-skewed as well, with a peak around 100 (the league average) and a small number of pitchers with an ERA+ above 200. The distributions of BB/9 and SO/9 are mostly symmetric, with peaks around 3 and 9, respectively.

```{r}

ip_sal = merged_pitching |> 
  ggplot(aes(x = ip_total, y = salary_2022)) + 
  geom_point(col = "red") + 
  scale_x_continuous(lim = c(0, 250)) +
  scale_y_continuous(labels = label_comma()) +
  labs(x = "Innings Pitched",
       y = "") +
  theme_bw()

era_sal = merged_pitching |> 
  ggplot(aes(x = era_2, y = salary_2022)) + 
  geom_point(col = "lightblue") +
  scale_x_continuous(lim = c(0, 400)) + 
  scale_y_continuous(labels = label_comma()) +
  labs(x = "ERA+",
       y = "") +
    guides(y="none") +
  theme_bw()

bb_sal = merged_pitching |> 
  ggplot(aes(x = bb9, y = salary_2022)) + 
  geom_point(col = "darkgreen") +
  scale_x_continuous(lim = c(0, 10)) + 
  scale_y_continuous(labels = label_comma()) +
   labs(x = "Walks per 9 Innings (BB/9)",
       y = "") +
  theme_bw()

so_sal = merged_pitching |> 
  ggplot(aes(x = so9, y = salary_2022)) + 
  geom_point(col = "orange") +
  scale_x_continuous(lim = c(0, 20)) + 
  scale_y_continuous(labels = label_comma()) +
   labs(x = "Strikeouts Per 9 Innings (SO/9)",
       y = "") +
    guides(y="none") +
  theme_bw()

scatt_pitch = (ip_sal + era_sal) / (bb_sal + so_sal)
scatt_pitch + plot_annotation(title = "Scatterplots of Pitching Statistics and Salary")

```

Interestingly, the pitchers with the highest ERA+ often had considerably lower salaries. 