---
title: "Pitching Model"
output:   
  html_document: 
    code_folding: hide
    toc: true
    toc_float: true
    df_print: kable

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rvest)
library(readxl)
library(scales)
library(patchwork)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%",
  dpi = 300,
  warning = FALSE,
  message = FALSE
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

salaries = read_excel(
  "data/MLB-Salaries 2000-24.xlsx", 
  sheet = "2022.xls",
  skip = 1) |>
  select(1:4) |>
  rename(position = "Pos'n",
         salary_2022 = "2022.0",
         service_time_yrs = "MLS",
         name = "Player") %>%
  mutate(name = str_split(name, ","),
         name = map(name, rev),       
         name = map_chr(name, str_c, collapse = " "),
         name = str_trim(name),
         name = str_replace_all(name, "[*#\\.,]", ""),
         junior = str_detect(name, "Jr"),
         name = if_else(junior == TRUE, str_remove(name, " Jr"), name),
         name = if_else(junior == TRUE, str_c(name, " Jr"), name),
         name = stringi::stri_trans_general(name,id = "Latin-ASCII")) |>
  select(-junior) %>%
  mutate(simple_position = str_split_i(position, "-", 1),
         simple_position = fct_relevel(simple_position,
                                       c("c", "1b", "2b", "3b", "ss",
                                         "lf", "cf", "rf", "inf", "of",
                                         "dh", "rhp", "lhp")),
         service_time_floor = floor(service_time_yrs),
         experience_group = case_when(service_time_floor <= 2 ~ "0-2 years",
                                      service_time_floor >= 3 ~ "3+ years"))

pitching = read_delim("data/2022 MLB Player Stats - Pitching.csv", delim = ";",
                     locale = locale(encoding = "latin1")) |>
  janitor::clean_names() %>%
  group_by(name) |>
  mutate(name_count = n(),
         keep_row = case_when(name_count == 1 ~ TRUE,
                              name_count > 1 & tm == "TOT" ~ TRUE,
                              .default = FALSE)) |>
  filter(keep_row == TRUE) %>%
  mutate(name_count = n(),
         keep_row = case_when(name_count == 1 ~ TRUE,
                              name_count > 1 & lg == "MLB" ~ TRUE,
                              .default = FALSE)) |>
  ungroup() |>
  filter(keep_row == TRUE) %>%
  mutate(name = str_split(name, "\\s+"),
         name = map_chr(name, str_c, collapse = " "),
         name = str_trim(name),
         name = str_replace_all(name, "[*#\\.]", ""),
         name = stringi::stri_trans_general(name,id = "Latin-ASCII"))

merged_pitching <- inner_join(salaries, pitching, by = "name") |>
  filter(ip >= 20) |>
  separate(ip, into = c("ip", "ip_dec"), remove = FALSE, convert = TRUE) |>
  mutate(ip_dec_333 = ifelse(is.na(ip_dec), 0, ip_dec * 333)  ,
         ip_total = paste(ip, ip_dec_333, sep = "."),
         ip_total = as.numeric(ip_total)) |>
  select(-ip, -ip_dec, -ip_dec_333) |>
  separate(position, into = c("hand", "pitcher_type")) |>
  mutate(pitcher_type = ifelse(is.na(pitcher_type), "r", pitcher_type)) |>
  filter(hand == "lhp" | hand == "rhp")

salary_not_in_pitching <- anti_join(salaries, pitching, by = "name")
pitching_not_in_salary <- anti_join(pitching, salaries, by = "name")
```

```{r}

lm_pitching = lm(log10(salary_2022) ~ hand + pitcher_type + hand*pitcher_type + ip_total + era_2 + bb9 + so9 + experience_group,
                 data = merged_pitching)

broom::tidy(lm_pitching)

broom::glance(lm_pitching)
```

```{r}

plot(residuals(lm_pitching) ~ fitted.values(lm_pitching),
     main = "Residuals vs. Fitted Values Plot for Pitching Model",
     xlab = "Fitted Values",
     ylab = "Residuals")
abline(h = 0)

```

```{r}

qqnorm(residuals(lm_pitching),
       main = "Normal Q-Q Plot for Pitching Model")
qqline(residuals(lm_pitching))

```

We decided to log-transform the salary variable because it is heavily right-skewed, and decided to use $log_{10}$ instead of $ln$ for the sake of interpretability. We included ERA+, the measure of a pitcher's ERA relative to the league average, as well as innings pitched (`ip_total`), walks per nine innings (`bb9`) and strikeouts per nine innings (`so9`); we used walks/strikeouts per nine innings since the total number of walks/strikeouts would be highly correlated with the number of innings pitched. 

We accounted for pitcher handedness (right vs. left) as well as pitcher type (the default type is a closing pitcher, with relief pitchers and starting pitchers also included in the model). Finally, we explored the interaction between handedness and pitcher type to see if pitcher type affects the relationship between handedness and salary.

From our model, we can see that pitcher handedness and pitcher type have no significant effect on expected salary. Interestingly, the `era_2` predictor, even when accounting for pitcher handedness, pitcher type, and the number of innings pitched, has an insignificant $P$-value, implying that there is no significant association between one of the most highly regarded summative statistics for pitching effectiveness and pitcher salary when accounting for pitcher handedness, pitcher type, innings pitched, walk and strikeout rates, and service time.

The residuals vs. fitted values plot shows that the residuals are generally evenly spread out; the straight line on the left side of the plot represents the large number of MLB players who are still on their rookie contracts, and therefore have a low fitted value (as they usually make between \$700,000 and \$1,000,000 per year, which is quite low by MLB standards). The Normal Q-Q plot shows that the model's residuals deviate only slightly from Normal behavior.

### Top 5 Pitchers by Value

```{r}

merged_pitching |>
  modelr::add_predictions(lm_pitching) |>
  modelr::add_residuals(lm_pitching) |>
  arrange(resid) |>
  select(name, salary_2022, service_time_floor, experience_group, era_2, pred, resid ) |>
  mutate(expected_salary = 10^pred) |>
  slice(1:5) 

```

### Top 10 Pitchers by Salary

```{r}

merged_pitching |>
  modelr::add_predictions(lm_pitching) |>
  modelr::add_residuals(lm_pitching) |>
  arrange(desc(salary_2022)) |>
  select(name, salary_2022, service_time_floor, experience_group, era_2, pred, resid ) |>
  mutate(expected_salary = 10^pred) |>
  slice(1:10) 

```
