---
title: "Exploratory Data Analysis"
output:   
  html_document: 
    code_folding: hide
    toc: true
    toc_float: true
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rvest)
library(readxl)
library(scales)
library(patchwork)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%",
  dpi = 300,
  warning = FALSE,
  message = FALSE
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```


## Salaries Data import and cleaning

Descrition of data cleaning: 

-   reading in data

-   selecting and renaming the columns

-   since names are in last, first format, reorder them to be first last

-   make sure all "Jr"s go at the end

-   remove accents using stringi::stri_trans_general

```{r}
salaries = read_excel(
  "data/MLB-Salaries 2000-24.xlsx", 
  sheet = "2022.xls",
  skip = 1) |>
  select(1:4) |>
  rename(position = "Pos'n",
         salary_2022 = "2022.0",
         service_time_yrs = "MLS",
         name = "Player") %>%
  mutate(name = str_split(name, ","),
         name = map(name, rev),       
         name = map_chr(name, str_c, collapse = " "),
         name = str_trim(name),
         name = str_replace_all(name, "[*#\\.,]", ""),
         junior = str_detect(name, "Jr"),
         name = if_else(junior == TRUE, str_remove(name, " Jr"), name),
         name = if_else(junior == TRUE, str_c(name, " Jr"), name),
         name = stringi::stri_trans_general(name,id = "Latin-ASCII")) |>
  select(-junior) %>%
  mutate(simple_position = str_split_i(position, "-", 1),
         simple_position = fct_relevel(simple_position,
                                       c("c", "1b", "2b", "3b", "ss",
                                         "lf", "cf", "rf", "inf", "of",
                                         "dh", "rhp", "lhp")),
         service_time_floor = floor(service_time_yrs))
```

## Batting and Pitching data import and cleaning

### Batting 
Descrition of data cleaning: 

-   reading in data and clean names

-   grouping by name and using a count to ensure we only end up with one of each name (slightly complicated due to interleague trades and potential for players with the same name)

-  remove whitespace, from names and use stringi again to remove accents so that the formatting exactly matches the salaries tibble

-   Filter out those with fewer than 100 plate appearances

```{r}
batting = read_delim("data/2022 MLB Player Stats - Batting.csv", delim = ";",
                     locale = locale(encoding = "latin1")) |>
  janitor::clean_names() %>%
  group_by(name) |>
  mutate(name_count = n(),
         keep_row = case_when(name_count == 1 ~ TRUE,
                              name_count > 1 & tm == "TOT" ~ TRUE,
                              .default = FALSE)) |>
  filter(keep_row == TRUE) %>%
  mutate(name_count = n(),
         keep_row = case_when(name_count == 1 ~ TRUE,
                              name_count > 1 & lg == "MLB" ~ TRUE,
                              .default = FALSE)) |>
  ungroup() |>
  filter(keep_row == TRUE) %>%
  mutate(name = str_split(name, "\\s+"),
         name = map_chr(name, str_c, collapse = " "),
         name = str_trim(name),
         name = str_replace_all(name, "[*#\\.]", ""),
         name = stringi::stri_trans_general(name,id = "Latin-ASCII"))

merged_batting <- salaries |>
  mutate(simple_position = str_split_i(position, "-", 1)) |>
  filter(!simple_position %in% c("rhp","lhp")) |>
  inner_join(batting, by = "name") |>
  filter(pa >= 100)

salary_not_in_batting <- anti_join(salaries, batting, by = "name")
batting_not_in_salary <- anti_join(batting, salaries, by = "name")
```

### Pitching

(same data cleaning process as above)
```{r}
pitching = read_delim("data/2022 MLB Player Stats - Pitching.csv", delim = ";",
                     locale = locale(encoding = "latin1")) |>
  janitor::clean_names() %>%
  group_by(name) |>
  mutate(name_count = n(),
         keep_row = case_when(name_count == 1 ~ TRUE,
                              name_count > 1 & tm == "TOT" ~ TRUE,
                              .default = FALSE)) |>
  filter(keep_row == TRUE) %>%
  mutate(name_count = n(),
         keep_row = case_when(name_count == 1 ~ TRUE,
                              name_count > 1 & lg == "MLB" ~ TRUE,
                              .default = FALSE)) |>
  ungroup() |>
  filter(keep_row == TRUE) %>%
  mutate(name = str_split(name, "\\s+"),
         name = map_chr(name, str_c, collapse = " "),
         name = str_trim(name),
         name = str_replace_all(name, "[*#\\.]", ""),
         name = stringi::stri_trans_general(name,id = "Latin-ASCII"))

merged_pitching <- inner_join(salaries, pitching, by = "name") |>
  filter(ip >= 20) |>
  separate(ip, into = c("ip", "ip_dec"), remove = FALSE, convert = TRUE) |>
  mutate(ip_dec_333 = ifelse(is.na(ip_dec), 0, ip_dec * 333)  ,
         ip_total = paste(ip, ip_dec_333, sep = "."),
         ip_total = as.numeric(ip_total)) |>
  select(-ip, -ip_dec, -ip_dec_333) |>
  separate(position, into = c("hand", "pitcher_type")) |>
  mutate(pitcher_type = ifelse(is.na(pitcher_type), "r", pitcher_type)) |>
  filter(hand == "lhp" | hand == "rhp")

salary_not_in_pitching <- anti_join(salaries, pitching, by = "name")
pitching_not_in_salary <- anti_join(pitching, salaries, by = "name")
```

Note that there are a number of names in `salaries` not found in `batting` or `pitching`, and vice versa. Some of this may be genuine missingness (e.g. salaries has fewer rows than pitching and batting combined), but some is also due to alternative spellings of names in the datasets. Possibly could be fixed using other matching methods, but the analysis pipeline will still work from here.

## Salaries EDA

Some Basic EDA for salaries on their own

Service time (i.e. number of years of experience)
```{r}
salaries |>
  count(service_time_floor) |>
  ggplot(aes(x = service_time_floor, y = n)) + 
  geom_col() + 
  labs(title = "Number of Players per Year of Experience",
       x = "Years in the MLB",
       y = "Number of Players") +
  scale_x_continuous(breaks = seq(0,21,by=1)) +
  theme_bw()
```
This is right skewed, so may need to transform it if used in regression

Salary
```{r}
salary_dist <- salaries |>
  ggplot(aes(x = salary_2022)) + 
  geom_density() + 
  labs(title = "2022 Salary Distribution",
       x = "2022 Salary",
       y = "") +
  scale_x_continuous(labels = label_comma()) +
  theme_bw()

salary_position_dist <- salaries |>
  mutate(position_group = case_match(simple_position,
                                     c("rhp","lhp") ~ "pitcher",
                                     c("cf","lf","rf","of") ~ "outfield",
                                     "dh" ~ "dh",
                                     .default = "infield")) |>
  ggplot(aes(x = salary_2022, color = position_group)) +
  geom_density() +
  labs(title = "Salary by Position Group",
       x = "2022 Salary",
       y = "") +
  scale_x_continuous(labels = label_comma()) +
  scale_color_discrete(name = "Position Group",
                       labels = c("Designated Hitter",
                                "Infield",
                                "Outfield",
                                "Pitcher")) +
  theme_bw()

salary_dist / salary_position_dist
```
This is also heavily right skewed, with many players having salary under $1,000,000

Position type count
```{r}
merged_batting |>
  count(simple_position) |> 
  mutate(simple_position = fct_infreq(simple_position,w = n)) |>
  ggplot(aes(x = simple_position, y = n)) +
  geom_col() +
  labs(title = "Players per Position",
       x = "",
       y = "Number of Players") +
  scale_x_discrete(guide = guide_axis(angle = 75), 
                  labels = c("1b" = "1st Base",
                    "2b" = "2nd Base",
                    "3b" = "3rd Base",
                    "inf" = "Infield",
                    "c" = "Catcher",
                    "rf" = "Right Field",
                    "lf" = "Left Field",
                    "cf" = "Center Field",
                    "of" = "Outfield",
                    "lhp" = "Left Hand Pitcher",
                    "rhp" = "Right Hand Pitcher",
                    "dh" = "Designated Hitter",
                    "ss" = "Shortstop")) +
  theme_bw()
```

salary vs experience, by position group
```{r}
salaries %>%
  mutate(position_group = case_match(simple_position,
                                     c("rhp","lhp") ~ "pitcher",
                                     c("cf","lf","rf","of") ~ "outfield",
                                     "dh" ~ "dh",
                                     .default = "infield")) %>%
  filter(service_time_floor<=10) %>%
  ggplot(aes(x=as.factor(service_time_floor),
             y=salary_2022, color=position_group)) +
  geom_point() + 
  facet_wrap(~position_group, nrow = 2) 

```

### Salaries by Position Boxplots

```{r}
salaries %>%
  ggplot(aes(x = simple_position, y = salary_2022)) +
  geom_boxplot() +
  labs(title = "Distribution of Salary by Position",
       x = "",
       y = "") +
  scale_x_discrete(guide = guide_axis(angle = 75), 
                  labels = c("1b" = "1st Base",
                    "2b" = "2nd Base",
                    "3b" = "3rd Base",
                    "inf" = "Infield",
                    "c" = "Catcher",
                    "rf" = "Right Field",
                    "lf" = "Left Field",
                    "cf" = "Center Field",
                    "of" = "Outfield",
                    "lhp" = "Left Hand Pitcher",
                    "rhp" = "Right Hand Pitcher",
                    "dh" = "Designated Hitter",
                    "ss" = "Shortstop")) +
  scale_y_continuous(labels = label_comma()) +
  theme_bw()
```

Experience by position
```{r}
salaries %>%
  ggplot(aes(x = simple_position, y = service_time_floor)) +
  geom_boxplot() +
  labs(title = "Distribution of Experience by Position",
       x = "",
       y = "") +
  scale_x_discrete(guide = guide_axis(angle = 75), 
                  labels = c("1b" = "1st Base",
                    "2b" = "2nd Base",
                    "3b" = "3rd Base",
                    "inf" = "Infield",
                    "c" = "Catcher",
                    "rf" = "Right Field",
                    "lf" = "Left Field",
                    "cf" = "Center Field",
                    "of" = "Outfield",
                    "lhp" = "Left Hand Pitcher",
                    "rhp" = "Right Hand Pitcher",
                    "dh" = "Designated Hitter",
                    "ss" = "Shortstop")) +
  scale_y_continuous(labels = label_comma()) +
  theme_bw()

```

### Salary Distribution by Experience

```{r}
salaries %>%
  ggplot(aes(x=factor(floor(service_time_yrs)), y=salary_2022)) +
  geom_boxplot() +
  labs(title = "Salary Distribution for Years of Experience",
       x = "Years of Experience",
       y = "") +
  scale_y_continuous(labels = label_comma()) +
  theme_bw()
```

### Positions by Experience

```{r}
salaries |>
  ggplot(aes(x = service_time_floor, 
             y = simple_position, 
             fill = simple_position)) + 
  ggridges::geom_density_ridges(alpha = 0.5) +
  labs(title = "Experience by Position",
       x = "Years of Experience",
       y = "") +
  xlim(0,25) +
  scale_y_discrete(labels = c("1b" = "1st Base",
                    "2b" = "2nd Base",
                    "3b" = "3rd Base",
                    "inf" = "Infield",
                    "c" = "Catcher",
                    "rf" = "Right Field",
                    "lf" = "Left Field",
                    "cf" = "Center Field",
                    "of" = "Outfield",
                    "lhp" = "Left Hand Pitcher",
                    "rhp" = "Right Hand Pitcher",
                    "dh" = "Designated Hitter",
                    "ss" = "Shortstop")) +
  scale_fill_discrete(name = "Position",
                      labels = c("1b" = "1st Base",
                    "2b" = "2nd Base",
                    "3b" = "3rd Base",
                    "inf" = "Infield",
                    "c" = "Catcher",
                    "rf" = "Right Field",
                    "lf" = "Left Field",
                    "cf" = "Center Field",
                    "of" = "Outfield",
                    "lhp" = "Left Hand Pitcher",
                    "rhp" = "Right Hand Pitcher",
                    "dh" = "Designated Hitter",
                    "ss" = "Shortstop")) +
  theme_bw()
  
```

## Batting EDA

### Distributions of variables

Looking at some hitting statistics
```{r}
merged_batting |> 
  ggplot(aes(x = hr)) + 
  geom_histogram() +
  labs(title = "Distribution of Home Runs",
    x = "Home Run Total",
    y = "Players")
```

```{r}
merged_batting |> 
  ggplot(aes(x = rbi)) + 
  geom_histogram() +
  labs(x = "Runs batted in (RBI)")
```

```{r}
merged_batting %>%
  ggplot(aes(x = ab,y= hr)) +
  geom_point(aes(color = ops)) +
  geom_smooth(method = lm, se = FALSE, color = "firebrick") +
  labs(title = "Who is the most 'effective' at the plate?",
       x = "Chances at Bat",
       y = "Number of Home Runs") +
  scale_color_continuous("OPS") +
  theme_bw()
  
```


OPS is an "all in one" statistic combining on-base percentage (OBP) and slugging (SLG).

OBP is calculated as $\frac{Hits (H) + Walks (BB) + Hit by pitch (HBP)}{At \ bats (AB) + Walks (BB) + sacrifice \ flies (SF) + Hit by pitch (HBP)}$ 

Slugging is calculated as $\frac{total \ bases (TB)}{At \ bats (AB)}$

OPS is the sum of these two statistics.

OPS+ (or adjusted OPS) is adjusted for the park and league averages.

```{r}
merged_batting |> 
  ggplot(aes(x = ops)) + 
  geom_histogram() +
  labs(x = "OPS")

merged_batting |> 
  ggplot(aes(x = ops_2)) + 
  geom_histogram() +
  labs(x = "OPS+")
```

One idea might be to compare OPS or OPS+ to more traditional statistics, like RBIs, HRs, or batting averages to OPS or OPS+ as single predictors of salary.

One other factor to consider is the correlation between some of these variables
```{r}
merged_batting |>
  select(hr, rbi, pa, ba, ops) |>
  GGally::ggpairs()
```

Some of these potential predictors are fairly strongly correlated (like RBI and HR, or BA and OPS), so it's important not to include too many collinear variables in a potential linear model.



### Relationship to salary

Looking at some relationships to salary

Position vs. salary
```{r}
merged_batting |>
  mutate(positions = str_split_i(position, "-",1))  |>
  ggplot(aes(x = positions,y= salary_2022)) +
  geom_boxplot()
```

```{r}
merged_batting |>
  mutate(positions = str_split_i(position, "-",1))  |>
  group_by(positions) |>
  summarize(avg_salary = mean(salary_2022, na.rm = TRUE),
            median_salary = median(salary_2022, na.rm = TRUE))
```

Some further ideas
```{r}
merged_batting %>%
  ggplot(aes(x = salary_2022, y = hr)) +
  geom_point(data=transform(merged_batting, simple_position=NULL),
             color = "grey", alpha = .5) +
  geom_point(color = "firebrick") +
  facet_wrap(vars(simple_position)) +
  theme_bw()
```

```{r}
yrs = merged_batting |>
  ggplot(aes(x = service_time_floor, y = salary_2022)) + 
  geom_point() +
  labs(x="Years of Experience",y="") +
  scale_y_continuous(labels = label_comma()) +
  theme_bw()
hrs = merged_batting |>
  ggplot(aes(x = hr, y = salary_2022)) + 
  geom_point() +
  guides(y="none") +
  labs(x="Home Runs",y="") +
  theme_bw()
bas = merged_batting |>
  ggplot(aes(x = ba, y = salary_2022)) + 
  geom_point() +
  labs(x="Batting Averages",y="") +
  scale_y_continuous(labels = label_comma()) +
  theme_bw()
ops = merged_batting |>
  ggplot(aes(x = ops_2, y = salary_2022)) + 
  geom_point() +
  guides(y="none") +
  labs(x="OPS+",y="") +
  theme_bw()

scatter_patch = (yrs+hrs)/(bas+ops)

scatter_patch + plot_annotation(title = "Scatterplot of Batting Statistics and Salary")
```

```{r}
yrsCol = merged_batting |>
  count(service_time_floor) |>
  ggplot(aes(x = service_time_floor, y=n)) + 
  geom_col() +
  labs(x="Years of Experience",y="") +
  theme_bw()
hrsHist = merged_batting |>
  ggplot(aes(x = hr)) + 
  geom_histogram() +
  labs(x="Home Runs",y="") +
  theme_bw()
basHist = merged_batting |>
  ggplot(aes(x = ba)) + 
  geom_histogram() +
  labs(x="Batting Averages",y="") +
  theme_bw()
opsHist = merged_batting |>
  ggplot(aes(x = ops_2)) + 
  geom_histogram() +
  labs(x="OPS+",y="") +
  theme_bw()

dist_patch = (yrsCol+hrsHist)/(basHist+opsHist)
dist_patch + plot_annotation(title = "Distribution of Important Batting Statistics")

```

Since salary (and other variables) are very skewed, we may want to transform these data using a log transformation.

### Regression models

Home runs
```{r}
hr_fit <- lm(log10(salary_2022) ~ hr,
             data = merged_batting)

hr_fit |>
  broom::tidy()
```


OPS+
```{r}
ops_fit <- lm(log10(salary_2022) ~ ops_2,
             data = merged_batting)

ops_fit |>
  broom::tidy()
```

Would also want to account for years of experience in predicting salary

```{r}
experience_fit <- lm(log10(salary_2022) ~ service_time_floor,
                     data = merged_batting)

experience_fit |>
  broom::tidy()
```

An interesting question might be: is OPS+ (an "all-in-one" statistic) a better predictor of salary than traditional statistics?

To do this, I'm going to account for batting average, home runs per plate appearance. I'm also going to adjust for service time for both models, since that is strongly correlated with salary.

```{r}
hr_ba_fit <- merged_batting |>
  mutate(hr_per_pa = hr / pa) |>
  lm(log10(salary_2022) ~ ba + hr_per_pa + service_time_floor,
                   data = _)

hr_ba_fit |>
  broom::tidy()

hr_ba_fit |>
  broom::glance()

plot(hr_ba_fit, which = 1)
plot(hr_ba_fit, which = 2)
```

```{r}
ops_fit <- merged_batting |>
  lm(log10(salary_2022) ~ ops_2 + service_time_floor,
                   data = _)

ops_fit |>
  broom::tidy()

ops_fit |>
  broom::glance()

plot(ops_fit, which = 1)
plot(ops_fit, which = 2)
```


Based on these statistics, the model with than OPS+ is a better predictor than the one with home runs and batting average, which makes sense since OPS+ is internally accounting for more other factors.

Which players provided the best value for money?
```{r}
merged_batting |>
  modelr::add_predictions(ops_fit) |>
  modelr::add_residuals(ops_fit) |>
  arrange(resid) |>
  select(name, salary_2022, service_time_floor, ops_2, pred, resid ) |>
  mutate(expected_salary = 10^pred) |>
  slice(1:10)
```

The most cost-effective players were those who

-   Played well (high OPS+) but had low salaries.

-   Played poorly and had a lot of experience, but had low salaries.

## Pitching EDA

In order to eliminate pitchers with very few appearances or position players (catchers, first basemen, etc.) who came in to pitch in blowouts, I filtered only for pitchers with at least 10 innings pitches.

```{r}

ip_plot = merged_pitching |> 
  ggplot(aes(x = ip_total)) + 
  geom_histogram(binwidth = 5, fill = "red", col = "black") +
  labs(
       x = "Innings Pitched",
       y = "Pitcher Count") +
  theme_bw()

era_plot = merged_pitching |>
  ggplot(aes(x = era_2)) + 
  geom_histogram(fill = "lightblue", col = "black") +
    labs(
       x = "ERA+",
       y = "Pitcher Count") +
  theme_bw()

bb_plot = merged_pitching |>
  ggplot(aes(x = bb9)) + 
  geom_histogram(fill = "darkgreen", col = "black") +
    labs(
       x = "Walks per 9 Innings (BB/9)",
       y = "Pitcher Count") +
  theme_bw()

so_plot = merged_pitching |>
  ggplot(aes(x = so9)) + 
  geom_histogram(fill = "orange", col = "black") +
    labs(
       x = "Strikeouts per 9 Innings (SO/9)",
       y = "Pitcher Count") +
  theme_bw()

hist_pitch = (ip_plot + era_plot) / (bb_plot + so_plot)
hist_pitch + plot_annotation(title = "Histograms of Pitching Statistics")

```

The distribution of innings pitched is quite right-skewed, with a peak around 70 innings pitched. The distribution of ERA+ is right-skewed as well, with a peak around 100 (the league average) and a small number of pitchers with an ERA+ above 200. The distributions of BB/9 and SO/9 are mostly symmetric, with peaks around 3 and 9, respectively.

```{r}

ip_sal = merged_pitching |> 
  ggplot(aes(x = ip_total, y = salary_2022)) + 
  geom_point(col = "red") + 
  scale_x_continuous(lim = c(0, 250)) +
  scale_y_continuous(labels = label_comma()) +
  labs(x = "Innings Pitched",
       y = "2022 Salary") +
  theme_bw()

era_sal = merged_pitching |> 
  ggplot(aes(x = era_2, y = salary_2022)) + 
  geom_point(col = "lightblue") +
  scale_x_continuous(lim = c(0, 400)) + 
  scale_y_continuous(labels = label_comma()) +
  labs(x = "ERA+",
       y = "2022 Salary") +
  theme_bw()

bb_sal = merged_pitching |> 
  ggplot(aes(x = bb9, y = salary_2022)) + 
  geom_point(col = "darkgreen") +
  scale_x_continuous(lim = c(0, 10)) + 
  scale_y_continuous(labels = label_comma()) +
   labs(x = "Walks per 9 Innings (BB/9)",
       y = "2022 Salary") +
  theme_bw()

so_sal = merged_pitching |> 
  ggplot(aes(x = so9, y = salary_2022)) + 
  geom_point(col = "orange") +
  scale_x_continuous(lim = c(0, 20)) + 
  scale_y_continuous(labels = label_comma()) +
   labs(x = "Strikeouts Per 9 Innings (SO/9)",
       y = "2022 Salary") +
  theme_bw()

scatt_pitch = (ip_sal + era_sal) / (bb_sal + so_sal)
scatt_pitch + plot_annotation(title = "Scatterplots of Pitching Statistics vs. Salary")

```


### IP vs. Log Salary Plot

```{r}

merged_pitching |> 
  ggplot(aes(x = ip_total, y = log(salary_2022))) + 
  geom_point() + 
  scale_x_continuous(lim = c(0, 400))

```

### ERA+ vs. Log Salary Plot

```{r}

merged_pitching |> 
  ggplot(aes(x = era_2, y = log(salary_2022))) + 
  geom_point() + 
  scale_x_continuous(lim = c(0, 400))

```

### LMs

```{r}

lm_ip = lm(salary_2022 ~ ip_total, data = merged_pitching)
summary(lm_ip)

lm_ip_data = tibble(fitted.values(lm_ip), residuals(lm_ip))

names(lm_ip_data) = c("fitted", "resid")

lm_ip_data |>
  ggplot(aes(x = fitted, y = resid)) +
  geom_point()

view(lm_ip_data)

```

```{r}

lm_era_2 = lm(salary_2022 ~ era_2, data = merged_pitching)
summary(lm_era_2)

lm_era_2_data = tibble(fitted.values(lm_era_2), residuals(lm_era_2))

names(lm_era_2_data) = c("fitted", "resid")

lm_era_2_data |>
  ggplot(aes(x = fitted, y = resid)) +
  geom_point()

view(lm_era_2_data)

```

### Log-Transformed LMs

```{r}

lm_ip_log = lm(log(salary_2022) ~ ip_total, data = merged_pitching)
summary(lm_ip_log)

lm_ip_log_data = tibble(fitted.values(lm_ip_log), residuals(lm_ip_log))

names(lm_ip_log_data) = c("fitted", "resid")

lm_ip_log_data |>
  ggplot(aes(x = fitted, y = resid)) +
  geom_point()

view(lm_ip_log_data)

```

```{r}

lm_era_2_log = lm(log(salary_2022) ~ era_2, data = merged_pitching)
summary(lm_era_2_log)

lm_era_2_data_log = tibble(fitted.values(lm_era_2_log), residuals(lm_era_2_log))

names(lm_era_2_data_log) = c("fitted", "resid")

lm_era_2_data_log |>
  ggplot(aes(x = fitted, y = resid)) +
  geom_point()

```

### Final Pitching LM

```{r}

lm_pitching = lm(salary_2022 ~ hand + pitcher_type + hand*pitcher_type + ip_total + era_2,
                 data = merged_pitching)

summary(lm_pitching)

```

From our model, we can see that relief pitchers have significantly lower salaries than closing pitchers, which makes sense, since closing pitchers have very strong reputations as some of the most effective pitchers in the game and therefore will draw high salaries. Additionally, the `era_2` predictor, even when accounting for pitcher handedness, pitcher type, and the number of innings pitched, has an insignificant $P$-value, implying that there is no significant association between one of the most highly regarded summative statistics for pitching effectiveness and pitcher salary.
